---
title: "PrimeSeq Gene Detection: Sense vs Sense+Antisense Comparison"
subtitle: "Comparing gene detection results with and without antisense intergenic reads"
author: "Felix Pförtner"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 3
    code_folding: hide
---

# Introduction

This analysis compares gene detection results between:
- **Sense-only data**: Normal gene expression counts (sense strand)
- **Sense+Antisense data**: Combined sense and antisense intergenic reads


# Setup

## Load Libraries

```{r setup-libraries}
library(tidyverse)
library(ggpubr)
library(here)
library(patchwork)  # For combining plots
```

## Initialize Environment

```{r setup-environment}
here::i_am("scripts/antisense_reads/Antisense_gene_detection/Detected_genes.Rmd")

source(here("functions/custom_functions_PS.R"))
```
## Define Parameters

```{r setup-parameters}
# Barcodes to exclude from analysis
unwanted_BCs <- c(
  "TTGTAAGCGGAA",
  "AACAGGTTCCAA",
  "GCTACTGGTCCA",
  "AACGTAACGCTG",
  "AGTCGTGTCCTG",
  "AATGTTCTCTCC",
  "GTCGCCAACTTC",
  "TGGCTGCAGGTT"
)

# Color scheme for plotting
color_vector <- c("original" = "#969994", # general protocols  
                  "improved" = "#96D8BF",   
                  "improved w/o PTO" = "#2C3A31",  
                  
                  "optimized \n DNase" = "#096581", # DNase test
                  "no DNase" = "#011E28",
                  
                  "P7-RS & P5+RS" = "#969994", # Seq adapters
                  "P7-RS & P5-RS" = "#d7c4c1",
                  "P7+RS & P5+RS" = "#8EDEE1", 
                  "P7+RS & P5-RS" = "#1C333F",
                  
                  "original" = "#969994", # Seq adapters
                  "blocked" = "#F0B0C9",
                  "mirrored" = "#490914",
                  
                  "50 °C" = "#3F1A4C", # RT temp
                  "42 °C" = "#DDC7E6",
                  
                  "80ng" = "#C4C082",
                  "320ng" = "#7D7A3B",
                  "920ng" = "#454321",
                  
                  "resuspended" = "#D28A4B" # Resuspension test (Suppl.)
)
```

# Data Loading and Processing

## Load Count Matrices

```{r load-data-paths}
old_paths <- c(here("data/FC2024_05_02_PoP96_BA/03_zUMIs/old/zUMIs_output/expression/OldNewPTO_old.dgecounts.rds"),
               here("data/FC2024_06_01_PoP96_FP_BA/03_zUMIs/old/zUMIs_output/expression/OldNewPTO_old.dgecounts.rds"),
               here("data/FC2024_08_01_PoP96III/03_zUMIs/old/zUMIs_output/expression/OldNewPTO_old.dgecounts.rds"),
               here("data/FC2024_11_01_PoP64_1/03_zUMIs/old/zUMIs_output/expression/PoP64_old.dgecounts.rds")
)

PTO_paths <- gsub("old", "PTO", old_paths)

# Project names for antisense files (matching order of old_paths)
antisense_project_names_old <- c("PoP96_1_old", "PoP96_2_old", "PoP96_3_old", "PoP64_old")
antisense_project_names_PTO <- c("PoP96_1_PTO", "PoP96_2_PTO", "PoP96_3_PTO", "PoP64_PTO")

# Paths to antisense count matrices
antisense_paths_old <- map_chr(antisense_project_names_old, ~ {
  here("scripts/antisense_reads/Antisense_power", .x, paste0(.x, "_antisense.dgecounts.rds"))
})

antisense_paths_PTO <- map_chr(antisense_project_names_PTO, ~ {
  here("scripts/antisense_reads/Antisense_power", .x, paste0(.x, "_antisense.dgecounts.rds"))
})

process_data <- function(paths, suffix, antisense_paths = NULL) {
  data_list <- map(seq_along(paths), ~ {
    data <- readRDS(paths[.x])$umicount$inex$all %>%
      as.matrix() %>%
      remove_Geneversion() %>%
      .[, !(colnames(.) %in% unwanted_BCs)]
    
    # Add antisense counts if provided
    if (!is.null(antisense_paths) && file.exists(antisense_paths[.x])) {
      antisense_data <- readRDS(antisense_paths[.x])$umicount$inex$all %>%
        as.matrix() %>%
        remove_Geneversion()
      
      # Align genes and barcodes
      all_genes <- union(rownames(data), rownames(antisense_data))
      all_barcodes <- union(colnames(data), colnames(antisense_data))
      
      # Create aligned matrices
      aligned_normal <- matrix(0, nrow = length(all_genes), ncol = length(all_barcodes))
      rownames(aligned_normal) <- all_genes
      colnames(aligned_normal) <- all_barcodes
      aligned_normal[rownames(data), colnames(data)] <- data
      
      aligned_antisense <- matrix(0, nrow = length(all_genes), ncol = length(all_barcodes))
      rownames(aligned_antisense) <- all_genes
      colnames(aligned_antisense) <- all_barcodes
      aligned_antisense[rownames(antisense_data), colnames(antisense_data)] <- antisense_data
      
      # Combine by adding counts
      data <- aligned_normal + aligned_antisense
      
      # Filter unwanted barcodes after combining
      data <- data[, !(colnames(data) %in% unwanted_BCs), drop = FALSE]
    }
    
    colnames(data) <- paste0(colnames(data), paste0("_", .x, "_", suffix))
    return(data)
  })
  
  # Create a union of all row names
  all_row_names <- Reduce(union, lapply(data_list, rownames))
  
  # Align each matrix to have the same row names, filling missing values with 0
  aligned_data <- map(data_list, ~ {
    aligned_matrix <- matrix(0, nrow = length(all_row_names), ncol = ncol(.x))
    rownames(aligned_matrix) <- all_row_names
    aligned_matrix[rownames(.x), ] <- .x
    colnames(aligned_matrix) <- colnames(.x)
    return(aligned_matrix)
  })
  
  # Combine all matrices into one wide matrix
  combined_matrix <- do.call(cbind, aligned_data)
  return(combined_matrix)
}

# Process original and improved data - SENSE ONLY (normal)
old_sense <- process_data(old_paths, "old")
PTO_sense <- process_data(PTO_paths, "PTO")
ncol(old_sense)
ncol(PTO_sense)

# Process original and improved data - SENSE + ANTISENSE (combined)
old_sense_antisense <- process_data(old_paths, "old", antisense_paths = antisense_paths_old)
PTO_sense_antisense <- process_data(PTO_paths, "PTO", antisense_paths = antisense_paths_PTO)
ncol(old_sense_antisense)
ncol(PTO_sense_antisense)

```

# Gene Detection Comparison

## Calculate Detected Genes per Sample

```{r calculate-detected-genes}
# Function to count detected genes per sample (genes with count > 0)
count_detected_genes <- function(count_matrix, threshold = 0) {
  detected <- colSums(count_matrix > threshold)
  return(detected)
}

# Calculate detected genes for each dataset
detected_old_sense <- count_detected_genes(old_sense)
detected_PTO_sense <- count_detected_genes(PTO_sense)
detected_old_sense_antisense <- count_detected_genes(old_sense_antisense)
detected_PTO_sense_antisense <- count_detected_genes(PTO_sense_antisense)

# Create summary data frame
detection_summary <- tibble(
  sample = c(names(detected_old_sense), names(detected_PTO_sense),
             names(detected_old_sense_antisense), names(detected_PTO_sense_antisense)),
  detected_genes = c(detected_old_sense, detected_PTO_sense,
                     detected_old_sense_antisense, detected_PTO_sense_antisense),
  condition = c(rep("original", length(detected_old_sense)),
                rep("improved", length(detected_PTO_sense)),
                rep("original", length(detected_old_sense_antisense)),
                rep("improved", length(detected_PTO_sense_antisense))),
  data_type = c(rep("Sense only", length(detected_old_sense) + length(detected_PTO_sense)),
                rep("Sense + Antisense", length(detected_old_sense_antisense) + length(detected_PTO_sense_antisense)))
) %>%
  mutate(
    condition = factor(condition, levels = c("original", "improved")),
    data_type = factor(data_type, levels = c("Sense only", "Sense + Antisense"))
  )

# Extract project number from sample name for grouping
detection_summary <- detection_summary %>%
  mutate(
    project = str_extract(sample, "_\\d+_") %>% 
      str_remove_all("_") %>%
      as.numeric() %>%
      replace_na(4)  # PoP64 samples don't have number, assign 4
  )

# Summary statistics
detection_stats <- detection_summary %>%
  group_by(condition, data_type) %>%
  summarise(
    n_samples = n(),
    mean_detected = mean(detected_genes),
    median_detected = median(detected_genes),
    sd_detected = sd(detected_genes),
    min_detected = min(detected_genes),
    max_detected = max(detected_genes),
    .groups = "drop"
  )

print(detection_stats)
```

## Visualizations

### Bar Plot: Mean Detected Genes by Condition

```{r barplot-mean-detected}
p1 <- detection_stats %>%
  ggplot(aes(x = condition, y = mean_detected, fill = data_type)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(aes(ymin = mean_detected - sd_detected, 
                    ymax = mean_detected + sd_detected),
                position = position_dodge(width = 0.8), width = 0.3) +
  theme_pubr() +
  labs(
    x = "Condition",
    y = "Mean Detected Genes",
    fill = "Data Type",
    title = "Mean Gene Detection: Sense vs Sense+Antisense"
  ) +
  theme(legend.position = "right")

p1
```

### Box Plot: Distribution of Detected Genes per Sample

```{r boxplot-detected-genes}
# Extract base sample info (same function as used later)
extract_base_info <- function(sample_name) {
  base <- str_remove(sample_name, "_(old|PTO)$")
  return(base)
}

# Prepare paired data for connecting lines and statistical tests
paired_data_wide <- detection_summary %>%
  mutate(base_sample = extract_base_info(sample)) %>%
  select(base_sample, condition, data_type, detected_genes) %>%
  pivot_wider(names_from = data_type, values_from = detected_genes) %>%
  filter(!is.na(`Sense only`) & !is.na(`Sense + Antisense`)) %>%
  mutate(
    `Sense only` = as.numeric(`Sense only`),
    `Sense + Antisense` = as.numeric(`Sense + Antisense`)
  )

# Convert to long format for plotting with connecting lines
paired_data <- paired_data_wide %>%
  pivot_longer(cols = c(`Sense only`, `Sense + Antisense`),
               names_to = "data_type", values_to = "detected_genes") %>%
  mutate(
    data_type = factor(data_type, levels = c("Sense only", "Sense + Antisense")),
    condition = factor(condition, levels = c("original", "improved"))
  )

# Perform paired statistical tests for each condition
stat_tests <- paired_data_wide %>%
  group_by(condition) %>%
  summarise(
    # Paired t-test
    t_test = list(t.test(`Sense only`, `Sense + Antisense`, paired = TRUE)),
    p_value = t_test[[1]]$p.value,
    # Wilcoxon signed-rank test (non-parametric alternative)
    wilcox_test = list(wilcox.test(`Sense only`, `Sense + Antisense`, paired = TRUE)),
    wilcox_p_value = wilcox_test[[1]]$p.value,
    .groups = "drop"
  ) %>%
  select(condition, p_value, wilcox_p_value)

print("Paired statistical tests:")
print(stat_tests)

# Create significance labels
stat_tests <- stat_tests %>%
  mutate(
    p_label = case_when(
      p_value < 0.0001 ~ "****",
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      TRUE ~ "ns"
    )
  )

# Calculate y positions for significance annotations
y_max <- max(detection_summary$detected_genes, na.rm = TRUE)
y_annot <- y_max * 1.05

# Define simple, distinct colors for sense vs sense+antisense
# Using colors that don't overlap with color_vector
sense_colors <- c("Sense only" = "#4A90E2",  # Nice blue
                  "Sense + Antisense" = "#FF6B6B")  # Nice coral

p2 <- detection_summary %>%
  ggplot(aes(x = condition, y = detected_genes, color = data_type)) +
  geom_boxplot(position = position_dodge(width = 0.8), width = 0.7, 
               outlier.alpha = 0.5) +
  geom_point(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
             alpha = 0.4, size = 1) +
  # Add significance annotations for each condition
  # Using annotation_custom or manual positioning with numeric x conversion
  geom_text(data = stat_tests,
            aes(x = condition, y = y_annot * 1.02, label = p_label),
            inherit.aes = FALSE, size = 4, vjust = 0) +
  scale_color_manual(values = sense_colors) +
  theme_pubr() +
  labs(
    x = "",
    y = "Detected Genes per Sample",
    color = ""
  ) +
  theme(legend.position = "bottom")+
  coord_flip()

# Create plot for genes only detected in sense+antisense per sample
# Function to get detected genes per sample
get_detected_genes_per_sample <- function(count_matrix, threshold = 0) {
  detected_list <- apply(count_matrix > threshold, 2, function(x) rownames(count_matrix)[x])
  return(detected_list)
}

# Get detected genes per sample for each dataset
detected_genes_old_sense <- get_detected_genes_per_sample(old_sense)
detected_genes_PTO_sense <- get_detected_genes_per_sample(PTO_sense)
detected_genes_old_sense_antisense <- get_detected_genes_per_sample(old_sense_antisense)
detected_genes_PTO_sense_antisense <- get_detected_genes_per_sample(PTO_sense_antisense)

# Calculate genes only in sense+antisense per sample
# For old/original condition - match samples by name
genes_only_per_sample_old <- map_dbl(names(detected_genes_old_sense_antisense), ~ {
  sample_name <- .x
  # Find matching sample in sense data (handle potential name mismatches)
  sense_sample <- names(detected_genes_old_sense)[names(detected_genes_old_sense) == sample_name]
  if (length(sense_sample) == 0) {
    # If no exact match, use all genes from sense+antisense
    sense_genes <- character(0)
  } else {
    sense_genes <- detected_genes_old_sense[[sense_sample]]
  }
  sense_antisense_genes <- detected_genes_old_sense_antisense[[sample_name]]
  only_antisense <- setdiff(sense_antisense_genes, sense_genes)
  length(only_antisense)
})

# For PTO/improved condition
genes_only_per_sample_PTO <- map_dbl(names(detected_genes_PTO_sense_antisense), ~ {
  sample_name <- .x
  # Find matching sample in sense data
  sense_sample <- names(detected_genes_PTO_sense)[names(detected_genes_PTO_sense) == sample_name]
  if (length(sense_sample) == 0) {
    sense_genes <- character(0)
  } else {
    sense_genes <- detected_genes_PTO_sense[[sense_sample]]
  }
  sense_antisense_genes <- detected_genes_PTO_sense_antisense[[sample_name]]
  only_antisense <- setdiff(sense_antisense_genes, sense_genes)
  length(only_antisense)
})

# Create per-sample data
genes_only_plot_data <- tibble(
  sample = c(names(detected_genes_old_sense_antisense), names(detected_genes_PTO_sense_antisense)),
  n_genes = c(genes_only_per_sample_old, genes_only_per_sample_PTO),
  condition = c(rep("original", length(genes_only_per_sample_old)),
                rep("improved", length(genes_only_per_sample_PTO)))
) %>%
  mutate(condition = factor(condition, levels = c("original", "improved")))

# Calculate summary statistics (mean and SEM)
genes_only_summary <- genes_only_plot_data %>%
  group_by(condition) %>%
  summarise(
    mean_genes = mean(n_genes),
    sem_genes = sd(n_genes) / sqrt(n()),
    .groups = "drop"
  )

p2b <- genes_only_plot_data %>%
  ggplot() +
  # Add points for each sample
  geom_point(aes(x = condition, y = n_genes), position = position_jitter(width = 0.2), alpha = 0.5, size = 1.5, 
             color = "gray40") +
  # Add bar for mean
  geom_col(data = genes_only_summary,
           aes(x = condition, y = mean_genes, fill = condition),
           width = 0.6, alpha = 0.7, inherit.aes = FALSE) +
  # Add SEM error bars
  geom_errorbar(data = genes_only_summary,
                aes(x = condition, y = mean_genes, ymin = mean_genes - sem_genes, 
                    ymax = mean_genes + sem_genes),
                width = 0.1, inherit.aes = FALSE) +
  scale_fill_manual(values = color_vector[c("original", "improved")]) +
  theme_pubr() +
  labs(
    x = "",
    y = "Genes Only Detected in Antisense"
  ) +
  theme(legend.position = "none")+
  coord_flip()

# Combine plots
plot <- (p2 | p2b) + plot_layout(widths = c(1, 0.8))
plot
saveRDS(p2, here("scripts/antisense_reads/Antisense_gene_detection/plot_abs.rds"))
saveRDS(p2b, here("scripts/antisense_reads/Antisense_gene_detection/plot_delta.rds"))
```

### Scatter Plot: Direct Comparison per Sample

```{r scatter-comparison}
# Match samples between sense and sense+antisense
# Extract base barcode and project info
# Sample names are: barcode_X_old or barcode_X_PTO where X is project number
extract_base_info <- function(sample_name) {
  # Remove only the suffix "_old" or "_PTO", keep project number
  # This ensures samples from different projects are kept separate
  base <- str_remove(sample_name, "_(old|PTO)$")
  return(base)
}

# Create comparison data
# First, check for any duplicates
comparison_data <- detection_summary %>%
  mutate(base_sample = extract_base_info(sample)) %>%
  select(base_sample, condition, data_type, detected_genes)

# Check for duplicates
duplicates <- comparison_data %>%
  count(base_sample, condition, data_type) %>%
  filter(n > 1)

if (nrow(duplicates) > 0) {
  cat("Warning: Found duplicate base_sample entries. Aggregating by taking mean.\n")
  comparison_data <- comparison_data %>%
    group_by(base_sample, condition, data_type) %>%
    summarise(detected_genes = mean(detected_genes), .groups = "drop")
}

# Now pivot
comparison_data <- comparison_data %>%
  pivot_wider(names_from = data_type, values_from = detected_genes) %>%
  filter(!is.na(`Sense only`) & !is.na(`Sense + Antisense`)) %>%
  # Ensure columns are numeric (in case of any list columns)
  mutate(
    `Sense only` = as.numeric(`Sense only`),
    `Sense + Antisense` = as.numeric(`Sense + Antisense`)
  )

p3 <- comparison_data %>%
  ggplot(aes(x = `Sense only`, y = `Sense + Antisense`, color = condition)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_abline(intercept = 0, slope = 1, linetype = 2, color = "gray50") +
  scale_color_manual(values = color_vector[c("original", "improved")]) +
  theme_pubr() +
  labs(
    x = "Detected Genes (Sense only)",
    y = "Detected Genes (Sense + Antisense)",
    color = "Condition",
    title = "Gene Detection: Sense vs Sense+Antisense"
  ) +
  theme(legend.position = "right")

p4 <- comparison_data %>%
  mutate(increase = `Sense + Antisense` - `Sense only`,
         percent_increase = (increase / `Sense only`) * 100) %>%
  ggplot(aes(x = condition, y = increase, fill = condition)) +
  geom_boxplot(width = 0.6, outlier.alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.2), alpha = 0.5) +
  scale_fill_manual(values = color_vector[c("original", "improved")]) +
  theme_pubr() +
  labs(
    x = "Condition",
    y = "Additional Genes Detected\n(Sense+Antisense - Sense only)",
    title = "Increase in Gene Detection"
  ) +
  theme(legend.position = "none")

# Combine plots
(p3 | p4) + plot_annotation(tag_levels = "A")
```



