---
title: "PrimeSeq Power Simulations: Sense vs Sense+Antisense Comparison"
subtitle: "Comparing power analysis results with and without antisense intergenic reads"
author: "Felix Pförtner"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 3
    code_folding: hide
---

# Introduction

This analysis compares power simulation results between:
- **Sense-only data**: Normal gene expression counts (sense strand)
- **Sense+Antisense data**: Combined sense and antisense intergenic reads

The goal is to assess how including antisense reads affects statistical power for differential expression analysis.

# Setup

## Load Libraries

```{r setup-libraries}
library(tidyverse)
library(powsimR)
library(ggpubr)
library(here)
library(patchwork)  # For combining plots
```

## Initialize Environment

```{r setup-environment}
here::i_am("scripts/antisense_reads/Antisense_power/Power_simulations.Rmd")

source(here("functions/custom_functions_PS.R"))

set.seed(925654)
```
## Define Parameters

```{r setup-parameters}
# Barcodes to exclude from analysis
unwanted_BCs <- c(
  "TTGTAAGCGGAA",
  "AACAGGTTCCAA",
  "GCTACTGGTCCA",
  "AACGTAACGCTG",
  "AGTCGTGTCCTG",
  "AATGTTCTCTCC",
  "GTCGCCAACTTC",
  "TGGCTGCAGGTT"
)

# Color scheme for plotting
color_vector <- c("original" = "#969994", # general protocols  
                  "improved" = "#96D8BF",   
                  "improved w/o PTO" = "#2C3A31",  
                  
                  "optimized \n DNase" = "#096581", # DNase test
                  "no DNase" = "#011E28",
                  
                  "P7-RS & P5+RS" = "#969994", # Seq adapters
                  "P7-RS & P5-RS" = "#d7c4c1",
                  "P7+RS & P5+RS" = "#8EDEE1", 
                  "P7+RS & P5-RS" = "#1C333F",
                  
                  "original" = "#969994", # Seq adapters
                  "blocked" = "#F0B0C9",
                  "mirrored" = "#490914",
                  
                  "50 °C" = "#3F1A4C", # RT temp
                  "42 °C" = "#DDC7E6",
                  
                  "80ng" = "#C4C082",
                  "320ng" = "#7D7A3B",
                  "920ng" = "#454321",
                  
                  "resuspended" = "#D28A4B" # Resuspension test (Suppl.)
)
```

# Data Loading and Processing

## Load Count Matrices

```{r load-data-paths}
old_paths <- c(here("data/FC2024_05_02_PoP96_BA/03_zUMIs/old/zUMIs_output/expression/OldNewPTO_old.dgecounts.rds"),
               here("data/FC2024_06_01_PoP96_FP_BA/03_zUMIs/old/zUMIs_output/expression/OldNewPTO_old.dgecounts.rds"),
               here("data/FC2024_08_01_PoP96III/03_zUMIs/old/zUMIs_output/expression/OldNewPTO_old.dgecounts.rds"),
               here("data/FC2024_11_01_PoP64_1/03_zUMIs/old/zUMIs_output/expression/PoP64_old.dgecounts.rds")
)

PTO_paths <- gsub("old", "PTO", old_paths)

# Project names for antisense files (matching order of old_paths)
antisense_project_names_old <- c("PoP96_1_old", "PoP96_2_old", "PoP96_3_old", "PoP64_old")
antisense_project_names_PTO <- c("PoP96_1_PTO", "PoP96_2_PTO", "PoP96_3_PTO", "PoP64_PTO")

# Paths to antisense count matrices
antisense_paths_old <- map_chr(antisense_project_names_old, ~ {
  here("scripts/antisense_reads/Antisense_power", .x, paste0(.x, "_antisense.dgecounts.rds"))
})

antisense_paths_PTO <- map_chr(antisense_project_names_PTO, ~ {
  here("scripts/antisense_reads/Antisense_power", .x, paste0(.x, "_antisense.dgecounts.rds"))
})

process_data <- function(paths, suffix, antisense_paths = NULL) {
  data_list <- map(seq_along(paths), ~ {
    data <- readRDS(paths[.x])$umicount$inex$all %>%
      as.matrix() %>%
      remove_Geneversion() %>%
      .[, !(colnames(.) %in% unwanted_BCs)]
    
    # Add antisense counts if provided
    if (!is.null(antisense_paths) && file.exists(antisense_paths[.x])) {
      antisense_data <- readRDS(antisense_paths[.x])$umicount$inex$all %>%
        as.matrix() %>%
        remove_Geneversion()
      
      # Align genes and barcodes
      all_genes <- union(rownames(data), rownames(antisense_data))
      all_barcodes <- union(colnames(data), colnames(antisense_data))
      
      # Create aligned matrices
      aligned_normal <- matrix(0, nrow = length(all_genes), ncol = length(all_barcodes))
      rownames(aligned_normal) <- all_genes
      colnames(aligned_normal) <- all_barcodes
      aligned_normal[rownames(data), colnames(data)] <- data
      
      aligned_antisense <- matrix(0, nrow = length(all_genes), ncol = length(all_barcodes))
      rownames(aligned_antisense) <- all_genes
      colnames(aligned_antisense) <- all_barcodes
      aligned_antisense[rownames(antisense_data), colnames(antisense_data)] <- antisense_data
      
      # Combine by adding counts
      data <- aligned_normal + aligned_antisense
      
      # Filter unwanted barcodes after combining
      data <- data[, !(colnames(data) %in% unwanted_BCs), drop = FALSE]
    }
    
    colnames(data) <- paste0(colnames(data), paste0("_", .x, "_", suffix))
    return(data)
  })
  
  # Create a union of all row names
  all_row_names <- Reduce(union, lapply(data_list, rownames))
  
  # Align each matrix to have the same row names, filling missing values with 0
  aligned_data <- map(data_list, ~ {
    aligned_matrix <- matrix(0, nrow = length(all_row_names), ncol = ncol(.x))
    rownames(aligned_matrix) <- all_row_names
    aligned_matrix[rownames(.x), ] <- .x
    colnames(aligned_matrix) <- colnames(.x)
    return(aligned_matrix)
  })
  
  # Combine all matrices into one wide matrix
  combined_matrix <- do.call(cbind, aligned_data)
  return(combined_matrix)
}

# Process old and PTO data - SENSE ONLY (normal)
old_sense <- process_data(old_paths, "old")
PTO_sense <- process_data(PTO_paths, "PTO")
ncol(old_sense)
ncol(PTO_sense)

# Process old and PTO data - SENSE + ANTISENSE (combined)
old_sense_antisense <- process_data(old_paths, "old", antisense_paths = antisense_paths_old)
PTO_sense_antisense <- process_data(PTO_paths, "PTO", antisense_paths = antisense_paths_PTO)
ncol(old_sense_antisense)
ncol(PTO_sense_antisense)

# Create batch information for sense-only data
batch_info_old_sense <- data.frame(batches = c(rep("a", 8), 
                                               rep("b", 8), 
                                               rep("c", 8), 
                                               rep("d", 8)
))
rownames(batch_info_old_sense) <- colnames(old_sense)

batch_info_PTO_sense <- data.frame(batches = c(rep("a", 8), 
                                               rep("b", 8), 
                                               rep("c", 8), 
                                               rep("d", 8)
))
rownames(batch_info_PTO_sense) <- colnames(PTO_sense)

# Create batch information for sense+antisense data
batch_info_old_sense_antisense <- data.frame(batches = c(rep("a", 8), 
                                                         rep("b", 8), 
                                                         rep("c", 8), 
                                                         rep("d", 8)
))
rownames(batch_info_old_sense_antisense) <- colnames(old_sense_antisense)

batch_info_PTO_sense_antisense <- data.frame(batches = c(rep("a", 8), 
                                                         rep("b", 8), 
                                                         rep("c", 8), 
                                                         rep("d", 8)
))
rownames(batch_info_PTO_sense_antisense) <- colnames(PTO_sense_antisense)
```

# PowsimR
## Evaluate Distribution
```{r}
# evalDistRes_old <- evaluateDist(countData = old,
#                             batchData = batch_info_old,
#                             RNAseq = "bulk", 
#                             Protocol = "UMI",
#                             Normalisation = "TMM",
#                             FracGenes = 0.1)
# plotEvalDist(evalDistRes_old)
# 
# 
# evalDistRes_old_MR <- evaluateDist(countData = old,
#                                 batchData = batch_info_old,
#                                 RNAseq = "bulk", 
#                                 Protocol = "UMI",
#                                 Normalisation = "MR",
#                                 FracGenes = 0.1)
# plotEvalDist(evalDistRes_old_MR)
# 
# evalDistRes_PTO <- evaluateDist(countData = PTO,
#                                 batchData = batch_info_PTO,
#                                 RNAseq = "bulk", 
#                                 Protocol = "UMI",
#                                 Normalisation = "TMM",
#                                 FracGenes = 0.01)
# plotEvalDist(evalDistRes_PTO)
```

## Estimate Parameters from Data

Estimate negative binomial parameters for both sense-only and sense+antisense datasets.

```{r estimate-parameters}
# SENSE ONLY
# estparam_old_sense <- estimateParam(countData = old_sense,
#                                     batchData = batch_info_old_sense,
#                                     Normalisation = "TMM",
#                                     RNAseq = "bulk",
#                                     Protocol = 'UMI',
#                                     Distribution = 'NB',
#                                     NCores = NULL,
#                                     verbose = TRUE,
#                                     GeneFilter = 0.1
# )


# estparam_PTO_sense <- estimateParam(countData = PTO_sense,
#                                     batchData = batch_info_PTO_sense,
#                                     Normalisation = "TMM",
#                                     RNAseq = "bulk",
#                                     Protocol = 'UMI',
#                                     Distribution = 'NB',
#                                     NCores = NULL,
#                                     verbose = TRUE,
#                                     GeneFilter = 0.1
# )


# SENSE + ANTISENSE
# estparam_old_sense_antisense <- estimateParam(countData = old_sense_antisense,
#                                               batchData = batch_info_old_sense_antisense,
#                                               Normalisation = "TMM",
#                                               RNAseq = "bulk",
#                                               Protocol = 'UMI',
#                                               Distribution = 'NB',
#                                               NCores = NULL,
#                                               verbose = TRUE,
#                                               GeneFilter = 0.1
# )


# estparam_PTO_sense_antisense <- estimateParam(countData = PTO_sense_antisense,
#                                               batchData = batch_info_PTO_sense_antisense,
#                                               Normalisation = "TMM",
#                                               RNAseq = "bulk",
#                                               Protocol = 'UMI',
#                                               Distribution = 'NB',
#                                               NCores = NULL,
#                                               verbose = TRUE,
#                                               GeneFilter = 0.1
# )


# plotParam(estParamRes = estparam_old_sense)
# plotParam(estParamRes = estparam_PTO_sense)
# plotParam(estParamRes = estparam_old_sense_antisense)
# plotParam(estParamRes = estparam_PTO_sense_antisense)
```

## Simulation Setup

```{r setup-simulation}
# # prime-seq
# p.lfc <- function(x) sample(c(-1.5,1.5), size=x,replace=T)*rgamma(x, shape = 1, rate = 2)
# # BRB-seq:
# lfc.gamma <- function(x) sample(c(-1,1), size=x,replace=T)*rgamma(x, 3, 3)
# # vignette:
# p.lfc <- function(x) sample(c(-1,1), size=x,replace=T)*rgamma(x, shape = 1, rate = 2)
# 
# # Setup simulations for SENSE ONLY data
# setupres_old_sense <- Setup(estParamRes = estparam_old_sense,
#                             nsims = 25,
#                             pLFC = p.lfc,
#                             n1 = c(3, 6, 12, 13, 14, 15, 16, 17, 18, 19, 20, 24, 48), 
#                             n2 = c(3, 6, 12, 13, 14, 15, 16, 17, 18, 19, 20, 24, 48), 
#                             setup.seed = 5327)
# 
# setupres_PTO_sense <- Setup(estParamRes = estparam_PTO_sense,
#                             nsims = 25,
#                             pLFC = p.lfc,
#                             n1 = c(3, 6, 12, 13, 14, 15, 16, 17, 18, 19, 20, 24, 48), 
#                             n2 = c(3, 6, 12, 13, 14, 15, 16, 17, 18, 19, 20, 24, 48), 
#                             setup.seed = 5327)
# 
# # Setup simulations for SENSE + ANTISENSE data
# setupres_old_sense_antisense <- Setup(estParamRes = estparam_old_sense_antisense,
#                                       nsims = 25,
#                                       pLFC = p.lfc,
#                                       n1 = c(3, 6, 12, 13, 14, 15, 16, 17, 18, 19, 20, 24, 48), 
#                                       n2 = c(3, 6, 12, 13, 14, 15, 16, 17, 18, 19, 20, 24, 48), 
#                                       setup.seed = 5327)
# 
# setupres_PTO_sense_antisense <- Setup(estParamRes = estparam_PTO_sense_antisense,
#                                       nsims = 25,
#                                       pLFC = p.lfc,
#                                       n1 = c(3, 6, 12, 13, 14, 15, 16, 17, 18, 19, 20, 24, 48), 
#                                       n2 = c(3, 6, 12, 13, 14, 15, 16, 17, 18, 19, 20, 24, 48), 
#                                       setup.seed = 5327)
```

## Run Simulations

**Note:** This step is computationally intensive and may take a long time.

```{r simulate-de}
# SENSE ONLY
# simres_old_sense <- simulateDE(SetupRes = setupres_old_sense,
#                                Normalisation = 'TMM',
#                                DEmethod = "limma-voom",
#                                verbose=T)
# 
# saveRDS(simres_old_sense, file = here("scripts/antisense_reads/Antisense_power/pow_simulated_old_sense.rds"))
simres_old_sense <- readRDS(here("scripts/antisense_reads/Antisense_power/pow_simulated_old_sense.rds"))
# 
# simres_PTO_sense <- simulateDE(SetupRes = setupres_PTO_sense,
#                                Normalisation = 'TMM',
#                                DEmethod = "limma-voom",
#                                verbose=T)
# 
# saveRDS(simres_PTO_sense, file = here("scripts/antisense_reads/Antisense_power/pow_simulated_PTO_sense.rds"))
simres_PTO_sense <- readRDS(here("scripts/antisense_reads/Antisense_power/pow_simulated_PTO_sense.rds"))
# 
# SENSE + ANTISENSE
# simres_old_sense_antisense <- simulateDE(SetupRes = setupres_old_sense_antisense,
#                                          Normalisation = 'TMM',
#                                          DEmethod = "limma-voom",
#                                          verbose=T)
# 
# saveRDS(simres_old_sense_antisense, file = here("scripts/antisense_reads/Antisense_power/pow_simulated_old_sense_antisense.rds"))
simres_old_sense_antisense <- readRDS(here("scripts/antisense_reads/Antisense_power/pow_simulated_old_sense_antisense.rds"))
# 
# simres_PTO_sense_antisense <- simulateDE(SetupRes = setupres_PTO_sense_antisense,
#                                          Normalisation = 'TMM',
#                                          DEmethod = "limma-voom",
#                                          verbose=T)
# 
# saveRDS(simres_PTO_sense_antisense, file = here("scripts/antisense_reads/Antisense_power/pow_simulated_PTO_sense_antisense.rds"))
simres_PTO_sense_antisense <- readRDS(here("scripts/antisense_reads/Antisense_power/pow_simulated_PTO_sense_antisense.rds"))
```

## Evaluate Differential Expression Results

```{r evaluate-de-results}
# Evaluate DE for SENSE ONLY data
evalderes_old_sense_mean <- evaluateDE(simRes = simres_old_sense,
                                       alpha.type = 'adjusted',
                                       MTC = 'BH',
                                       stratify.by = 'mean',
                                       filter.by = 'mean',
                                       strata.filtered = 1,
                                       target.by = 'lfc',
                                       delta = 1
)

evalderes_PTO_sense_mean <- evaluateDE(simRes = simres_PTO_sense,
                                       alpha.type = 'adjusted',
                                       MTC = 'BH',
                                       stratify.by = 'mean',
                                       filter.by = 'mean',
                                       strata.filtered = 1,
                                       target.by = 'lfc',
                                       delta = 1
)

evalderes_old_sense_lfc <- evaluateDE2(simRes = simres_old_sense,
                                       alpha.type = 'adjusted',
                                       MTC = 'BH',
                                       stratify.by = 'lfc_abs',
                                       filter.by = 'mean',
                                       strata.filtered = 1,
                                       target.by = 'lfc'
)

evalderes_PTO_sense_lfc <- evaluateDE2(simRes = simres_PTO_sense,
                                       alpha.type = 'adjusted',
                                       MTC = 'BH',
                                       stratify.by = 'lfc_abs',
                                       filter.by = 'mean',
                                       strata.filtered = 1,
                                       target.by = 'lfc'
)

# Evaluate DE for SENSE + ANTISENSE data
evalderes_old_sense_antisense_mean <- evaluateDE(simRes = simres_old_sense_antisense,
                                                 alpha.type = 'adjusted',
                                                 MTC = 'BH',
                                                 stratify.by = 'mean',
                                                 filter.by = 'mean',
                                                 strata.filtered = 1,
                                                 target.by = 'lfc',
                                                 delta = 1
)

evalderes_PTO_sense_antisense_mean <- evaluateDE(simRes = simres_PTO_sense_antisense,
                                                 alpha.type = 'adjusted',
                                                 MTC = 'BH',
                                                 stratify.by = 'mean',
                                                 filter.by = 'mean',
                                                 strata.filtered = 1,
                                                 target.by = 'lfc',
                                                 delta = 1
)

evalderes_old_sense_antisense_lfc <- evaluateDE2(simRes = simres_old_sense_antisense,
                                                 alpha.type = 'adjusted',
                                                 MTC = 'BH',
                                                 stratify.by = 'lfc_abs',
                                                 filter.by = 'mean',
                                                 strata.filtered = 1,
                                                 target.by = 'lfc'
)

evalderes_PTO_sense_antisense_lfc <- evaluateDE2(simRes = simres_PTO_sense_antisense,
                                                 alpha.type = 'adjusted',
                                                 MTC = 'BH',
                                                 stratify.by = 'lfc_abs',
                                                 filter.by = 'mean',
                                                 strata.filtered = 1,
                                                 target.by = 'lfc'
)


```

# Results Visualization

## Diagnostic Plots

```{r diagnostic-plots}
# SENSE ONLY
plotEvalDE(evalRes = evalderes_old_sense_mean, 
           rate = 'marginal', 
           quick = TRUE, 
           Annot = TRUE)
plotEvalDE(evalRes = evalderes_PTO_sense_mean, 
           rate = 'marginal', 
           quick = TRUE, 
           Annot = TRUE)

# SENSE + ANTISENSE
plotEvalDE(evalRes = evalderes_old_sense_antisense_mean, 
           rate = 'marginal', 
           quick = TRUE, 
           Annot = TRUE)
plotEvalDE(evalRes = evalderes_PTO_sense_antisense_mean, 
           rate = 'marginal', 
           quick = TRUE, 
           Annot = TRUE)

plotEvalDE(evalRes = evalderes_old_sense_lfc, rate = 'marginal', quick = TRUE, Annot = TRUE)
plotEvalDE(evalRes = evalderes_PTO_sense_lfc, rate = 'marginal', quick = TRUE, Annot = TRUE)
plotEvalDE(evalRes = evalderes_old_sense_antisense_lfc, rate = 'marginal', quick = TRUE, Annot = TRUE)
plotEvalDE(evalRes = evalderes_PTO_sense_antisense_lfc, rate = 'marginal', quick = TRUE, Annot = TRUE)
```

## Marginal Power Analysis

Power analysis stratified by sample size.

```{r marginal-power-analysis}
# Prepare data for SENSE ONLY
df_mean_sense <- getEvalDE_df_marginal(evalderes_old_sense_mean, 
                                       method = "original",
                                       count_type = "inex")
df_mean_sense <- bind_rows(df_mean_sense,
                           getEvalDE_df_marginal(evalderes_PTO_sense_mean, 
                                                 method = "improved",
                                                 count_type = "inex")) %>%
  mutate(Var1 = factor( str_extract(Var1, 
                                    "\\d*"),
                        levels = c("3", "6", "12", "24", "48"))) %>%
  filter(Var1 %in% as.character(c(3:48))) %>%
  mutate(data_type = "sense_only")

# Prepare data for SENSE + ANTISENSE
df_mean_sense_antisense <- getEvalDE_df_marginal(evalderes_old_sense_antisense_mean, 
                                                 method = "original",
                                                 count_type = "inex")
df_mean_sense_antisense <- bind_rows(df_mean_sense_antisense,
                                     getEvalDE_df_marginal(evalderes_PTO_sense_antisense_mean, 
                                                           method = "improved",
                                                           count_type = "inex")) %>%
  mutate(Var1 = factor( str_extract(Var1, 
                                    "\\d*"),
                        levels = c("3", "6", "12", "24", "48"))) %>%
  filter(Var1 %in% as.character(c(3:48))) %>%
  mutate(data_type = "sense_antisense")

# Combine both
df_mean <- bind_rows(df_mean_sense, df_mean_sense_antisense)%>%
  mutate(method = factor(method, levels=c("original", "improved")))

refval <- data.frame(L1 = c("FDR", "TPR"),
                     ref = c(0.1, 0.8))

power_overview_plot <- ggplot(data = df_mean %>% 
                                filter(L1 %in% c("TPR", "FDR")) %>%
                                mutate(L1 = factor(L1, levels=c("FDR", "TPR"))), 
                              aes(x =factor(Var1),
                                  y = value,
                                  color = method,
                                  linetype = data_type)) +
  stat_summary(fun.data = "mean_se",
               size = 0.3,) +
  stat_summary(aes(group = paste(method, data_type)),
               fun = mean, geom="line")+
  labs(x = "Samples per group", y = "Rate", color = "", linetype = "") +
  geom_hline(data = refval,
             aes(yintercept = ref),
             linetype="dashed",
             color='black')+
  theme_pubr(legend = "bottom")+
  facet_grid(L1~., scales = "free")+
  scale_color_manual(values = color_vector)+
  scale_linetype_manual(values=c(sense_only="solid", sense_antisense="dotted"))
```

## Conditional Power Analysis by Log2 Fold Change

Power analysis stratified by absolute log2 fold change.

```{r conditional-power-lfc}
# Prepare conditional data for SENSE ONLY
df_rates_lfc_sense <- getEvalDE_list_conditional(evalderes_old_sense_lfc,
                                                 method = "original",
                                                 count_type = "inex")$dat.stratified.long

df_rates_lfc_sense <- bind_rows(df_rates_lfc_sense,getEvalDE_list_conditional(evalderes_PTO_sense_lfc,
                                                                              method = "improved",
                                                                              count_type = "inex")$dat.stratified.long) %>%
  mutate(Var1 = str_replace(Var1, "0.82", "0.8")) %>%
  mutate(Var1 = str_replace(Var1, "0.81", "0.8")) %>%
  mutate(Var1 = str_replace(Var1, "1.16", "1.17")) %>%
  mutate(Var1 = str_replace(Var1, "0.25", "0.26")) %>%
  mutate(Var1 = str_replace(Var1, "0.61", "0.6")) %>%
  mutate(Var2 = factor( str_extract(Var2, 
                                    "\\d*"),
                        levels = c("3", "6", "12", "24", "48")))%>%
  filter(Var2 %in% as.character(c(3:48))) %>%
  mutate(data_type = "sense_only")

# Prepare conditional data for SENSE + ANTISENSE
df_rates_lfc_sense_antisense <- getEvalDE_list_conditional(evalderes_old_sense_antisense_lfc,
                                                           method = "original",
                                                           count_type = "inex")$dat.stratified.long

df_rates_lfc_sense_antisense <- bind_rows(df_rates_lfc_sense_antisense,getEvalDE_list_conditional(evalderes_PTO_sense_antisense_lfc,
                                                                                                  method = "improved",
                                                                                                  count_type = "inex")$dat.stratified.long) %>%
  mutate(Var1 = str_replace(Var1, "1.15", "1.17")) %>%
  mutate(Var1 = str_replace(Var1, "0.82", "0.8")) %>%
  mutate(Var1 = str_replace(Var1, "0.81", "0.8")) %>%
  mutate(Var1 = str_replace(Var1, "0.25", "0.26")) %>%
  mutate(Var1 = str_replace(Var1, "0.61", "0.6")) %>%
  mutate(Var2 = factor( str_extract(Var2, 
                                    "\\d*"),
                        levels = c("3", "6", "12", "24", "48")))%>%
  filter(Var2 %in% as.character(c(3:48))) %>%
  mutate(data_type = "sense_antisense")

# Combine both
df_rates_lfc <- bind_rows(df_rates_lfc_sense, df_rates_lfc_sense_antisense) %>%
  mutate(method = factor(method, levels=c("original", "improved")))

sample_cols<-c("#264653","#2a9d8f","#e9c46a","#f4a261","#e76f51")


power_LFC_plot <- ggplot(data = df_rates_lfc%>% 
                           filter(L1 %in% c("TPR", "FDR")) %>%
                           mutate(L1 = factor(L1, levels=c("FDR", "TPR"))), 
                         aes(y=value, 
                             x=Var1, 
                             #linetype=method, 
                             color=Var2,
                             linetype = data_type))+
  stat_summary(aes(x = Var1,
                   y = value,
                   group = paste(method, Var2, data_type)
  ),
  fun = mean, geom="line") +
  facet_grid(L1~method, scales = "free")+
  labs(x="Stratum abs. Log2 Fold Change",
       y="Rate",
       colour="Samples",
       linetype=""
  )+
  theme_pubr(legend = "none")+
  theme(axis.text.x=element_text(angle=60,hjust=1))+
  scale_color_manual(values = sample_cols)+
  scale_linetype_manual(values=c(sense_only="solid", sense_antisense="dotted"))+
  geom_hline(data = refval,
             aes(yintercept = ref),
             linetype="dashed",
             color='black')
```

## Conditional Power Analysis by Mean Expression

Power analysis stratified by mean expression level.

```{r conditional-power-mean}
# Prepare mean expression stratified data for SENSE ONLY
df_rates_mean_sense <- getEvalDE_list_conditional(evalderes_old_sense_mean,
                                                  method = "original",
                                                  count_type = "inex")$dat.stratified.long

df_rates_mean_sense <- bind_rows(df_rates_mean_sense,getEvalDE_list_conditional(evalderes_PTO_sense_mean,
                                                                                method = "improved",
                                                                                count_type = "inex")$dat.stratified.long) %>%
  mutate(data_type = "sense_only")

df_rates_mean_sense <- df_rates_mean_sense %>% 
  group_by(method) %>% 
  mutate(Var1=case_when(Var1==unique(Var1)[1]~0.1,
                        Var1==unique(Var1)[2]~0.2,
                        Var1==unique(Var1)[3]~0.3,
                        Var1==unique(Var1)[4]~0.4,
                        Var1==unique(Var1)[5]~0.5,
                        Var1==unique(Var1)[6]~0.6,
                        Var1==unique(Var1)[7]~0.7,
                        Var1==unique(Var1)[8]~0.8,
                        Var1==unique(Var1)[9]~0.9,
                        Var1==unique(Var1)[10]~1)) %>%
  mutate(Var2 = factor( str_extract(Var2, 
                                    "\\d*"),
                        levels = c("3", "6", "12", "24", "48")))%>%
  filter(Var2 %in% as.character(c(3:48)))

# Prepare mean expression stratified data for SENSE + ANTISENSE
df_rates_mean_sense_antisense <- getEvalDE_list_conditional(evalderes_old_sense_antisense_mean,
                                                            method = "original",
                                                            count_type = "inex")$dat.stratified.long

df_rates_mean_sense_antisense <- bind_rows(df_rates_mean_sense_antisense,getEvalDE_list_conditional(evalderes_PTO_sense_antisense_mean,
                                                                                                    method = "improved",
                                                                                                    count_type = "inex")$dat.stratified.long) %>%
  mutate(data_type = "sense_antisense")

df_rates_mean_sense_antisense <- df_rates_mean_sense_antisense %>% 
  group_by(method) %>% 
  mutate(Var1=case_when(Var1==unique(Var1)[1]~0.1,
                        Var1==unique(Var1)[2]~0.2,
                        Var1==unique(Var1)[3]~0.3,
                        Var1==unique(Var1)[4]~0.4,
                        Var1==unique(Var1)[5]~0.5,
                        Var1==unique(Var1)[6]~0.6,
                        Var1==unique(Var1)[7]~0.7,
                        Var1==unique(Var1)[8]~0.8,
                        Var1==unique(Var1)[9]~0.9,
                        Var1==unique(Var1)[10]~1)) %>%
  mutate(Var2 = factor( str_extract(Var2, 
                                    "\\d*"),
                        levels = c("3", "6", "12", "24", "48")))%>%
  filter(Var2 %in% as.character(c(3:48)))

# Combine both
df_rates_mean <- bind_rows(df_rates_mean_sense, df_rates_mean_sense_antisense)%>%
  mutate(method = factor(method, levels=c("original", "improved")))

power_expr_plot <- ggplot(data = df_rates_mean %>%
                            filter(L1 %in% c("TPR", "FDR")) %>%
                            mutate(L1 = factor(L1, levels=c("FDR", "TPR"))), 
                          aes(y=value, x=Var1, linetype=data_type, color=Var2))+
  stat_summary(aes(x = Var1,
                   y = value,
                   group = paste(method, Var2, data_type)),
               fun = mean, 
               geom="line") +
  facet_grid(L1~method, 
             scales = "free")+
  labs(x="Percentile Log2 Mean Expression",
       y="",
       colour="n",
       linetype="")+
  theme_pubr()+
  ylim(c(0,1))+
  theme(axis.text.x=element_text(angle=60,
                                 hjust=1),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank())+
  scale_color_manual(values = sample_cols)+
  scale_linetype_manual(values=c(sense_only="solid", sense_antisense="dotted"))+
  geom_hline(data = refval,
             aes(yintercept = ref),
             linetype="dashed",
             color='black')
```

## Combined Power Analysis Figure

```{r combine-power-plots}
# load other plots showing gene detection
p2 <- readRDS(here("scripts/antisense_reads/Antisense_gene_detection/plot_abs.rds"))
p2b <- readRDS(here("scripts/antisense_reads/Antisense_gene_detection/plot_delta.rds"))


design <- c("AABCCC\nDDDEEE")
design <- c("ACD\nBCD\nEEF")

legend <- as_ggplot(ggpubr::get_legend(power_expr_plot))

power_expr_plot <- power_expr_plot + theme(legend.position = "none")

power_plots_a <- (p2 + labs(tag = "A")) + 
  (p2b + labs(tag = "B")) + 
  plot_spacer()+
  (power_overview_plot + labs(tag = "C")) +
  (power_LFC_plot + labs(tag = "D")) + 
  (power_expr_plot + labs(tag = "E")) + 
  plot_layout(design = design,
              heights = c(0.6, 0.3,1),
              widths = c(0.85, 0.15, 1))

power_plots <- power_plots_a / legend + plot_layout(heights = c(1, 0.07))

power_plots

ggsave(power_plots,
       filename=here("figures/fig_S8_power_detection_antisense.pdf"),
       height=9,
       width=12
)
```



# Mean-Dispersion and Mean-Variance Relationships

## Extract Mean-Dispersion Data

```{r extract-mean-dispersion-data}
# Extract mean-dispersion data from all parameter estimates
mean_disp_old_sense <- Get_Mean_Disp(estparam_old_sense)
mean_disp_PTO_sense <- Get_Mean_Disp(estparam_PTO_sense)
mean_disp_old_sense_antisense <- Get_Mean_Disp(estparam_old_sense_antisense)
mean_disp_PTO_sense_antisense <- Get_Mean_Disp(estparam_PTO_sense_antisense)

# Combine data points
meanvsdisp.dat <- bind_rows(
  mean_disp_old_sense$meanvsdisp.dat %>% mutate(method = "original", data_type = "sense_only"),
  mean_disp_PTO_sense$meanvsdisp.dat %>% mutate(method = "improved", data_type = "sense_only"),
  mean_disp_old_sense_antisense$meanvsdisp.dat %>% mutate(method = "original", data_type = "sense_antisense"),
  mean_disp_PTO_sense_antisense$meanvsdisp.dat %>% mutate(method = "improved", data_type = "sense_antisense")
) %>%
  mutate(method = factor(method, levels = c("original", "improved")),
         data_type = factor(data_type, levels = c("sense_only", "sense_antisense")))

# Combine fitted data
meanvsdisp.fdat <- bind_rows(
  mean_disp_old_sense$meanvsdisp.fdat %>% mutate(method = "original", data_type = "sense_only"),
  mean_disp_PTO_sense$meanvsdisp.fdat %>% mutate(method = "improved", data_type = "sense_only"),
  mean_disp_old_sense_antisense$meanvsdisp.fdat %>% mutate(method = "original", data_type = "sense_antisense"),
  mean_disp_PTO_sense_antisense$meanvsdisp.fdat %>% mutate(method = "improved", data_type = "sense_antisense")
) %>%
  mutate(method = factor(method, levels = c("original", "improved")),
         data_type = factor(data_type, levels = c("sense_only", "sense_antisense")))

# Common dispersion values
cdisp_df <- data.frame(
  method = c("original", "improved", "original", "improved"),
  data_type = c("sense_only", "sense_only", "sense_antisense", "sense_antisense"),
  cdisp = c(mean_disp_old_sense$cdisp,
            mean_disp_PTO_sense$cdisp,
            mean_disp_old_sense_antisense$cdisp,
            mean_disp_PTO_sense_antisense$cdisp)
) %>%
  mutate(method = factor(method, levels = c("original", "improved")),
         data_type = factor(data_type, levels = c("sense_only", "sense_antisense")))

# Create overlay plot comparing sense vs sense+antisense for each method
meanvsdisp.plot_overlay <- ggplot() +
  # Data points - sense only (lighter)
  geom_point(data = meanvsdisp.dat %>% filter(data_type == "sense_only"),
             aes(x = Mean, y = Dispersion, color = "Sense only"),
             size = 0.2, alpha = 0.2) +
  # Data points - sense + antisense (darker)
  geom_point(data = meanvsdisp.dat %>% filter(data_type == "sense_antisense"),
             aes(x = Mean, y = Dispersion, color = "Sense + Antisense"),
             size = 0.2, alpha = 0.2) +
  # Fitted lines - sense only
  geom_line(data = meanvsdisp.fdat %>% filter(data_type == "sense_only"),
            aes(x = Mean, y = Dispersion, linetype = "Sense only", color = "Sense only"),
            linewidth = 1.2) +
  geom_line(data = meanvsdisp.fdat %>% filter(data_type == "sense_only"),
            aes(x = Mean, y = Lower, linetype = "Sense only", color = "Sense only"),
            linewidth = 0.8, alpha = 0.5) +
  geom_line(data = meanvsdisp.fdat %>% filter(data_type == "sense_only"),
            aes(x = Mean, y = Upper, linetype = "Sense only", color = "Sense only"),
            linewidth = 0.8, alpha = 0.5) +
  # Fitted lines - sense + antisense
  geom_line(data = meanvsdisp.fdat %>% filter(data_type == "sense_antisense"),
            aes(x = Mean, y = Dispersion, linetype = "Sense + Antisense", color = "Sense + Antisense"),
            linewidth = 1.2) +
  geom_line(data = meanvsdisp.fdat %>% filter(data_type == "sense_antisense"),
            aes(x = Mean, y = Lower, linetype = "Sense + Antisense", color = "Sense + Antisense"),
            linewidth = 0.8, alpha = 0.5) +
  geom_line(data = meanvsdisp.fdat %>% filter(data_type == "sense_antisense"),
            aes(x = Mean, y = Upper, linetype = "Sense + Antisense", color = "Sense + Antisense"),
            linewidth = 0.8, alpha = 0.5) +
  # Common dispersion lines
  geom_hline(data = cdisp_df,
             aes(yintercept = cdisp, linetype = data_type, color = data_type),
             linewidth = 0.8, alpha = 0.6) +
  labs(y = "Log Dispersion",
       x = "Log Mean",
       color = "Data type",
       linetype = "Data type") +
  scale_color_manual(values = c("Sense only" = "#2E86AB", 
                                "Sense + Antisense" = "#A23B72",
                                "sense_only" = "#2E86AB",
                                "sense_antisense" = "#A23B72"),
                     labels = c("Sense only" = "Sense only",
                                "Sense + Antisense" = "Sense + Antisense",
                                "sense_only" = "Sense only",
                                "sense_antisense" = "Sense + Antisense")) +
  scale_linetype_manual(values = c("Sense only" = 2,
                                   "Sense + Antisense" = 1,
                                   "sense_only" = 2,
                                   "sense_antisense" = 1),
                        labels = c("Sense only" = "Sense only",
                                   "Sense + Antisense" = "Sense + Antisense",
                                   "sense_only" = "Sense only",
                                   "sense_antisense" = "Sense + Antisense")) +
  facet_wrap(~ method,
             labeller = labeller(method = c(original = "Original", 
                                            improved = "Improved"))) +
  theme_pubr() +
  theme(legend.position = "right",
        strip.background = element_blank(),
        strip.text = element_text(size = 11, face = "bold"))

meanvsdisp.plot_overlay
```

## Mean-Dispersion Lines Only

Cleaner view with only fitted lines and confidence bands.

```{r mean-dispersion-lines}
meanvsdisp.plot_lines <- ggplot() +
  # Fitted lines - sense only
  geom_line(data = meanvsdisp.fdat %>% filter(data_type == "sense_only"),
            aes(x = Mean, y = Dispersion, linetype = "Sense only", color = "Sense only"),
            linewidth = 1.5) +
  geom_ribbon(data = meanvsdisp.fdat %>% filter(data_type == "sense_only"),
              aes(x = Mean, ymin = Lower, ymax = Upper, fill = "Sense only"),
              alpha = 0.2) +
  # Fitted lines - sense + antisense
  geom_line(data = meanvsdisp.fdat %>% filter(data_type == "sense_antisense"),
            aes(x = Mean, y = Dispersion, linetype = "Sense + Antisense", color = "Sense + Antisense"),
            linewidth = 1.5) +
  geom_ribbon(data = meanvsdisp.fdat %>% filter(data_type == "sense_antisense"),
              aes(x = Mean, ymin = Lower, ymax = Upper, fill = "Sense + Antisense"),
              alpha = 0.2) +
  # Common dispersion lines
  geom_hline(data = cdisp_df,
             aes(yintercept = cdisp, linetype = data_type, color = data_type),
             linewidth = 0.8, alpha = 0.6) +
  labs(y = "Log Dispersion",
       x = "Log Mean",
       color = "Data type",
       linetype = "Data type",
       fill = "Data type") +
  scale_color_manual(values = c("Sense only" = "#2E86AB", 
                                "Sense + Antisense" = "#A23B72",
                                "sense_only" = "#2E86AB",
                                "sense_antisense" = "#A23B72"),
                     labels = c("Sense only" = "Sense only",
                                "Sense + Antisense" = "Sense + Antisense",
                                "sense_only" = "Sense only",
                                "sense_antisense" = "Sense + Antisense")) +
  scale_fill_manual(values = c("Sense only" = "#2E86AB", 
                               "Sense + Antisense" = "#A23B72"),
                    labels = c("Sense only" = "Sense only",
                               "Sense + Antisense" = "Sense + Antisense")) +
  scale_linetype_manual(values = c("Sense only" = 2,
                                   "Sense + Antisense" = 1,
                                   "sense_only" = 2,
                                   "sense_antisense" = 1),
                        labels = c("Sense only" = "Sense only",
                                   "Sense + Antisense" = "Sense + Antisense",
                                   "sense_only" = "Sense only",
                                   "sense_antisense" = "Sense + Antisense")) +
  facet_wrap(~ method,
             labeller = labeller(method = c(original = "Original", 
                                            improved = "Improved"))) +
  theme_pubr() +
  theme(legend.position = "right",
        strip.background = element_blank(),
        strip.text = element_text(size = 11, face = "bold"))

meanvsdisp.plot_lines
```


