---
title: "RT genes"
output:
  html_document:
    df_print: paged
---

# Background

Reviewer comment: The finding of "extra" genes detected with the original RTase is interesting. 
Are these "extra" genes "real" or stochastic "noise"? 
It would be helpful to know if they are all lowly expressed, if they are reproducibly expressed across replicates, and if they are associated with particular biological pathways (e.g., via gene set analysis).

# Setup
```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(data.table)
library(parallel)
library(AnnotationDbi)
library(org.Mm.eg.db)
library(patchwork)
library(scales)
library(forcats)
library(here)
library(ggpubr)
library(clusterProfiler)
```

```{r setup_here}
here::i_am("scripts/RT_genes.Rmd")
```

# Data Loading

## Read UMI count data
```{r read_counts}
counts_M42 <- as.data.frame(as.matrix(readRDS(here("data/RTase/counts/Maxima_42.dgecounts.rds"))$umicount$inex$all))
counts_M50 <- as.data.frame(as.matrix(readRDS(here("data/RTase/counts/Maxima_50.dgecounts.rds"))$umicount$inex$all))
counts_S42 <- as.data.frame(as.matrix(readRDS(here("data/RTase/counts/SS_42.dgecounts.rds"))$umicount$inex$all))
counts_S50 <- as.data.frame(as.matrix(readRDS(here("data/RTase/counts/SS_50.dgecounts.rds"))$umicount$inex$all))
```

## Library size summary

```{r library_size_summary}
# Calculate mean and minimum library sizes per sample for each condition
mean_reads_M42 <- mean(colSums(counts_M42))
mean_reads_M50 <- mean(colSums(counts_M50))
mean_reads_S42 <- mean(colSums(counts_S42))
mean_reads_S50 <- mean(colSums(counts_S50))

min_reads_M42 <- min(colSums(counts_M42))
min_reads_M50 <- min(colSums(counts_M50))
min_reads_S42 <- min(colSums(counts_S42))
min_reads_S50 <- min(colSums(counts_S50))

avg_reads_df <- data.frame(
  matrix = c("Maxima_42", "Maxima_50", "SS_42", "SS_50"),
  avg_reads = c(mean_reads_M42, mean_reads_M50, mean_reads_S42, mean_reads_S50),
  min_reads = c(min_reads_M42, min_reads_M50, min_reads_S42, min_reads_S50)
)
avg_reads_df
```

# Data Normalization

To enable fair comparison across conditions, we need to downsample to a common depth. 
To minimize downsampling, we drop the lowest-depth replicate from each condition, keeping 3 replicates per condition.

## Downsampling and replicate selection
```{r downsample_counts}
set.seed(92384598)  # make the random downsampling reproducible

# Helper functions
as_int_mat <- function(df){
  m <- as.matrix(df)
  storage.mode(m) <- "integer"
  rownames(m) <- sub("\\.\\d+$", "", rownames(m))  # drop Ensembl version if present
  m
}

align_to_union_genes <- function(mats){
  all_genes <- Reduce(union, lapply(mats, rownames))
  lapply(mats, function(m){
    out <- matrix(0L, nrow = length(all_genes), ncol = ncol(m),
                  dimnames = list(all_genes, colnames(m)))
    out[rownames(m), ] <- m
    out
  })
}

drop_lowest_rep <- function(m, label){
  lib <- colSums(m)
  drop <- names(which.min(lib))
  message(sprintf("[%s] drop lowest: %s (UMIs=%d)", label, drop, lib[drop]))
  m[, setdiff(colnames(m), drop), drop = FALSE]
}

downsample_to <- function(m, depth){
  ds_col <- function(x){
    tot <- sum(x)
    if (tot < depth) stop("A kept sample has fewer UMIs than target depth.")
    if (tot == depth) return(x)
    p <- x / tot
    as.integer(rmultinom(1, size = depth, prob = p)[,1])
  }
  out <- apply(m, 2, ds_col)
  # apply() may drop dimnames depending on edge cases
  out <- as.matrix(out)
  storage.mode(out) <- "integer"
  rownames(out) <- rownames(m)
  colnames(out) <- colnames(m)
  out
}

# Prepare and align genes across all 4 conditions
mats <- list(
  M42 = as_int_mat(counts_M42),
  M50 = as_int_mat(counts_M50),
  S42 = as_int_mat(counts_S42),
  S50 = as_int_mat(counts_S50)
)
mats <- align_to_union_genes(mats)

# Keep top 3 replicates per condition (drop lowest-depth replicate)
mats_k <- list(
  M42 = drop_lowest_rep(mats$M42, "Maxima_42"),
  M50 = drop_lowest_rep(mats$M50, "Maxima_50"),
  S42 = drop_lowest_rep(mats$S42, "SS_42"),
  S50 = drop_lowest_rep(mats$S50, "SS_50")
)

# Choose one common downsample depth for all 12 kept samples
all_kept_libsizes <- unlist(lapply(mats_k, colSums))
target_depth <- min(all_kept_libsizes)
message(sprintf("Downsampling ALL kept samples to target_depth = %d UMIs", target_depth))

# Downsample each kept sample to target_depth
mats_ds <- lapply(mats_k, downsample_to, depth = target_depth)

# Output: downsampled count matrices (genes x 3 reps each)
# All columns now have identical total UMIs = target_depth
counts_M42_ds <- mats_ds$M42
counts_M50_ds <- mats_ds$M50
counts_S42_ds <- mats_ds$S42
counts_S50_ds <- mats_ds$S50

# Sanity check: verify all columns sum to target_depth
sapply(mats_ds, colSums)


```
# Gene Expression Analysis

## Count expressed genes per condition

```{r count_expressed_genes}
# For each count matrix, count in how many samples each gene is expressed (>0)
expressed_in_M42 <- rowSums(counts_M42_ds > 0)
expressed_in_M50 <- rowSums(counts_M50_ds > 0)
expressed_in_S42 <- rowSums(counts_S42_ds > 0)
expressed_in_S50 <- rowSums(counts_S50_ds > 0)

# Combine results into one data frame
# Get the union of all gene names across all conditions
all_genes <- Reduce(union, list(rownames(counts_M42), rownames(counts_M50_ds), rownames(counts_S42_ds), rownames(counts_S50_ds)))

# Helper function to align expressed counts to the union of all genes
get_expressed <- function(genes, expressed_vec) {
  expressed_out <- rep(0, length(all_genes))
  names(expressed_out) <- all_genes
  expressed_out[genes] <- expressed_vec
  return(expressed_out)
}

expressed_df <- data.frame(
  gene = all_genes,
  expressed_in_M42 = get_expressed(rownames(counts_M42_ds), expressed_in_M42),
  expressed_in_M50 = get_expressed(rownames(counts_M50_ds), expressed_in_M50),
  expressed_in_S42 = get_expressed(rownames(counts_S42_ds), expressed_in_S42),
  expressed_in_S50 = get_expressed(rownames(counts_S50_ds), expressed_in_S50)
)


```


## UpSet plot of gene expression overlap (not included)

```{r upset_plot, message=FALSE, warning=FALSE}
library(UpSetR)

# Define genes expressed in at least 2 samples for each condition
genes_M42 <- expressed_df$gene[expressed_df$expressed_in_M42 >= 2]
genes_M50 <- expressed_df$gene[expressed_df$expressed_in_M50 >= 2]
genes_S42 <- expressed_df$gene[expressed_df$expressed_in_S42 >= 2]
genes_S50 <- expressed_df$gene[expressed_df$expressed_in_S50 >= 2]

# Create a list of gene sets
gene_sets <- list(
  M42 = genes_M42,
  M50 = genes_M50,
  S42 = genes_S42,
  S50 = genes_S50
)

# Create upset plot
upset(fromList(gene_sets),
      sets = c("M42", "M50", "S42", "S50"),
      order.by = "freq",
      sets.bar.color = "#56B4E9",
      matrix.color = "#56B4E9",
      main.bar.color = "gray23",
      text.scale = c(1.3, 1.3, 1, 1, 1.5, 1))
```


## Mean-variance relationship (not included)

```{r mean_variance_plot}
# Calculate detection, mean, and variance statistics for each condition
gene_stats <- function(mat, prefix){
  tibble(
    gene = rownames(mat),
    !!paste0("det_", prefix) := rowSums(mat > 0),
    !!paste0("mean_", prefix) := rowMeans(mat),
    !!paste0("var_", prefix) := apply(mat, 1, var)
  )
}

st <- gene_stats(counts_M42_ds, "M42") %>%
  left_join(gene_stats(counts_M50_ds, "M50"), by="gene") %>%
  left_join(gene_stats(counts_S42_ds, "S42"), by="gene") %>%
  left_join(gene_stats(counts_S50_ds, "S50"), by="gene") %>%
  mutate(
    # With 3 kept reps: robust = detected in all 3; absent = detected in <=1
    robust_M42 = det_M42 >= 3 & mean_M42 >= 2,
    absent_S42 = det_S42 <= 1,
    absent_M50 = det_M50 <= 1,
    absent_S50 = det_S50 <= 1,
    
    extra_rtase = robust_M42 & absent_S42,
    extra_temp  = robust_M42 & absent_M50,
    extra_only_M42 = robust_M42 & absent_S42 & absent_M50 & absent_S50,
    
    category = case_when(
      extra_only_M42 ~ "only in M42",
      extra_rtase & extra_temp ~ "lost by RTase+temp",
      extra_rtase ~ "lost by RTase",
      extra_temp ~ "lost by temp",
      TRUE ~ "other"
    )
  )

ggplot(st, aes(x = mean_M42 + 1, y = var_M42 + 1, color = category)) +
  scale_x_log10() + scale_y_log10() +
  geom_point(alpha = 0.55, size = 1.1) +
  geom_abline(intercept = 0, slope = 1, linetype = 2) +  # Poisson reference line (var = mean)
  geom_smooth()+
  theme_classic() +
  labs(x="Mean UMI in M42 (+1, log10)", y="Variance across M42 reps (+1, log10)",
       title="Mean–variance in M42; genes 'lost' after RTase/temp change highlighted")

```



# Expression Level Analysis

## Expression levels of "lost" genes

```{r expression_levels}
# Using depth-matched, top 3 replicate matrices (genes x reps):
# counts_M42_ds, counts_M50_ds, counts_S42_ds, counts_S50_ds

det_n <- function(mat, thr = 0) rowSums(mat > thr)

# Calculate mean CPM across replicates for a matrix (genes x reps)
mean_cpm <- function(mat){
  lib <- colSums(mat)
  cpm <- sweep(mat, 2, lib / 1e6, "/")  # CPM per replicate
  rowMeans(cpm)
}

# Per-gene summary
df <- tibble(
  gene = rownames(counts_M42_ds),
  
  cpm_M42 = mean_cpm(counts_M42_ds),
  
  det_M42 = det_n(counts_M42_ds, thr = 0),
  det_M50 = det_n(counts_M50_ds, thr = 0),
  det_S42 = det_n(counts_S42_ds, thr = 0),
  det_S50 = det_n(counts_S50_ds, thr = 0)
)

# Define gene groups
robust_ref <- 2  # Out of 3 reps: reproducible in M42
lost_thr   <- 1  # Out of 3 reps: absent/rare in comparison

df2 <- df %>%
  mutate(
    expressed_all = det_M42 >= robust_ref & det_M50 >= robust_ref & det_S42 >= robust_ref & det_S50 >= robust_ref,
    
    lost_S42 = det_M42 >= robust_ref & det_S42 <= lost_thr,
    lost_M50  = det_M42 >= robust_ref & det_M50 <= lost_thr,
    lost_S50  = det_M42 >= robust_ref & det_S50 <= lost_thr,
    
    logCPM = log10(cpm_M42 + 1)
  ) %>%
  # Pivot to long format so each gene can appear in multiple categories
  pivot_longer(
    cols = c(expressed_all, lost_S50, lost_S42, lost_M50),
    names_to = "category_flag",
    values_to = "in_category"
  ) %>%
  filter(in_category) %>%
  mutate(
    group = case_when(
      category_flag == "expressed_all" ~ "expressed in all conditions",
      category_flag == "lost_S50" ~ "not in Superscript IV 50 °C",
      category_flag == "lost_S42" ~ "not in Superscript IV 42 °C",
      category_flag == "lost_M50" ~ "not in Maxima H- 50 °C",
      TRUE ~ NA_character_
    )
  ) %>%
  select(-category_flag, -in_category) %>%
  filter(!is.na(group))

# Create boxplot with proper ordering
group_levels <- c(
  "not in Superscript IV 50 °C",
  "not in Superscript IV 42 °C",
  "not in Maxima H- 50 °C",
  "expressed in all conditions"
)

df2p <- df2 %>%
  mutate(group = factor(group, levels = group_levels)) %>%
  filter(!is.na(group))


plot_expression <- ggplot(df2p, aes(x = group, y = logCPM)) +
  geom_boxplot(width = 0.65, outlier.alpha = 0.12, linewidth = 0.4) +
  coord_flip() +
  theme_pubr() +
  theme(
    legend.position = "none",
    axis.title.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "grey90")
  ) +
  labs(
    y = "log10(CPM + 1) in Maxima H- 42 °C"
  )

plot_expression
```

**Interpretation:** Genes that are lost upon switching RTase and/or temperature are predominantly lowly expressed in the reference condition (M42). This is consistent with many 'extra' detections arising close to the detection limit rather than representing a distinct highly expressed gene set.

## Reproducibility of "lost" genes
```{r reproducibility_analysis}
# Helper function to count detections
det_n <- function(mat, thr = 0) rowSums(mat > thr)

# Per-gene detection counts
df_det <- tibble(
  gene = rownames(counts_M42_ds),
  det_M42 = det_n(counts_M42_ds, thr = 0),
  det_M50 = det_n(counts_M50_ds, thr = 0),
  det_S42 = det_n(counts_S42_ds, thr = 0),
  det_S50 = det_n(counts_S50_ds, thr = 0)
)

nrep <- ncol(counts_M42_ds)

# Use same group definitions as before
robust_ref <- ceiling(2/3 * nrep)  # For 3 reps -> 2
lost_thr   <- floor(1/3 * nrep)    # For 3 reps -> 1

df_groups <- df_det %>%
  mutate(
    expressed_all = det_M42 >= robust_ref & det_M50 >= robust_ref & det_S42 >= robust_ref & det_S50 >= robust_ref,
    
    lost_rtase = det_M42 >= robust_ref & det_S42 <= lost_thr,
    lost_temp  = det_M42 >= robust_ref & det_M50 <= lost_thr,
    lost_both  = det_M42 >= robust_ref & det_S50 <= lost_thr,
    
    det_frac_M42 = det_M42 / nrep
  ) %>%
  # Pivot to long format so each gene can appear in multiple categories
  pivot_longer(
    cols = c(expressed_all, lost_both, lost_rtase, lost_temp),
    names_to = "category_flag",
    values_to = "in_category"
  ) %>%
  filter(in_category) %>%
  mutate(
    group = case_when(
      category_flag == "expressed_all" ~ "expressed in all conditions",
      category_flag == "lost_both" ~ "not in Superscript IV 50 °C",
      category_flag == "lost_rtase" ~ "not in Superscript IV 42 °C",
      category_flag == "lost_temp" ~ "not in Maxima H- 50 °C",
      TRUE ~ NA_character_
    )
  ) %>%
  select(-category_flag, -in_category) %>%
  filter(!is.na(group))

# Create reproducibility plot

nrep <- ncol(counts_M42_ds)

group_levels <- c(
  "not in Superscript IV 50 °C",
  "not in Superscript IV 42 °C",
  "not in Maxima H- 50 °C",
  "expressed in all conditions"
)

pdat <- df_groups %>%
  mutate(
    group = factor(group, levels = group_levels),
    det_class = ifelse(det_M42 == nrep,
                       "Yes",
                       "No")
  ) %>%
  dplyr::count(group, det_class) %>%
  group_by(group) %>%
  mutate(frac = n / sum(n)) %>%
  ungroup() %>%
  group_by(group) %>%
  mutate(label = ifelse(det_class == "Yes",
                        percent(frac, accuracy = 1), "")) %>%  # Label only one segment
  ungroup()

plot_reproducibility <- ggplot(pdat, aes(x = group, y = frac, fill = det_class)) +
  geom_col(width = 0.8, color = "white", linewidth = 0.4) +
  geom_text(aes(label = label),
            position = position_stack(vjust = 0.5),
            size = 3.6, color = "white") +
  scale_y_continuous(labels = percent_format(accuracy = 1), expand = expansion(mult = c(0, 0.02))) +
  scale_fill_manual(values = c(
    "No" = "#E76F51",
    "Yes" = "#2A9D8F"
  )) +
  coord_flip() +
  theme_pubr(legend = "right")+
  labs(
    x="",
    y = "Fraction of genes",
    fill = "Expression in all \nMaxima H- 42°C \nsamples?"
  )+
  theme( axis.text.y = element_blank(),
         axis.ticks.y = element_blank())

plot_reproducibility

```

## Combine expression and reproducibility plots

```{r combine_expression_plots}
plot_comb1 <- 
  plot_expression + 
  plot_reproducibility + 
  plot_layout(
    widths = c(1.3, 1))

plot_comb1

```

# Gene Set Enrichment Analysis

Are the 'lost' genes associated with any biological process?

## Preparation
```{r gsea_preparation}
strip_ver <- function(x) sub("\\.\\d+$", "", x)

lost_groups <- c("not in Superscript IV 50 °C", "not in Superscript IV 42 °C", "not in Maxima H- 50 °C")

lost_lists <- df2p %>%
  filter(group %in% lost_groups) %>%
  mutate(gene = as.character(gene),
         group = as.character(group)) %>%
  group_by(group) %>%
  summarise(genes = list(unique(gene)), .groups = "drop") %>%
  deframe()

# Inspect sizes (this will reveal any empty groups)
sapply(lost_lists, length)

universe_genes <- df2p %>% pull(gene) %>% as.character() %>% unique()

# Safe conversion to Entrez IDs
to_entrez_safe <- function(ids){
  # Coerce to plain character vector
  ids <- unlist(ids, use.names = FALSE)
  ids <- as.character(ids)
  ids <- strip_ver(ids)
  ids <- ids[!is.na(ids) & ids != ""]
  ids <- unique(ids)
  
  if (length(ids) == 0) return(character(0))  # Prevents error for empty input
  
  eg <- mapIds(org.Mm.eg.db,
               keys = ids,
               keytype = "ENSEMBL",
               column = "ENTREZID",
               multiVals = "first")
  unique(na.omit(unname(eg)))
}

lost_lists_ez <- lapply(lost_lists, to_entrez_safe)
universe_ez   <- to_entrez_safe(universe_genes)

# Sanity check: how many genes mapped?
sapply(lost_lists_ez, length)
length(universe_ez)
```

## Run GO Biological Process enrichment
```{r go_enrichment}
run_go <- function(gene_ez, universe_ez){
  enrichGO(
    gene          = gene_ez,
    universe      = universe_ez,
    OrgDb         = org.Mm.eg.db,
    keyType       = "ENTREZID",
    ont           = "BP",
    pAdjustMethod = "BH",
    pvalueCutoff  = 0.05,
    qvalueCutoff  = 0.2,
    readable      = TRUE
  )
}

ego <- lapply(lost_lists_ez, run_go, universe_ez = universe_ez)

# Create a compact results table
ego_tbl <- bind_rows(lapply(names(ego), function(nm){
  x <- ego[[nm]]
  if (is.null(x) || nrow(as.data.frame(x)) == 0) return(NULL)
  as.data.frame(x) %>% mutate(list = nm)
}))

# Helper function to convert ratio strings to numeric values
ratio_to_num <- function(r){
  # Input format: "12/345"
  r <- as.character(r)
  out <- rep(NA_real_, length(r))
  ok <- !is.na(r) & grepl("/", r)
  parts <- strsplit(r[ok], "/", fixed = TRUE)
  out[ok] <- vapply(parts, function(z) as.numeric(z[1]) / as.numeric(z[2]), numeric(1))
  out
}

ego_fe_tbl <- imap_dfr(ego, function(res, nm){
  if (is.null(res)) return(NULL)
  df <- as.data.frame(res)
  if (nrow(df) == 0) return(NULL)
  
  df %>%
    mutate(
      category = nm,
      GeneRatio_num = ratio_to_num(GeneRatio),
      BgRatio_num   = ratio_to_num(BgRatio),
      FoldEnrichment = GeneRatio_num / BgRatio_num
    )
})

# Keep only significant terms
sig <- ego_fe_tbl %>%
  filter(!is.na(FoldEnrichment), p.adjust < 0.05)

# Pick top N terms per category by fold enrichment
top_n <- 10
plot_df <- sig %>%
  group_by(category) %>%
  slice_max(order_by = FoldEnrichment, n = top_n, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    term = paste0(category, ": ", Description),
    term = fct_reorder(term, FoldEnrichment)
  )%>%
  mutate(Description_wrapped = str_wrap(Description, width = 55))

# plot
enrichment_plot <- ggplot(plot_df, aes(x = FoldEnrichment, y = Description_wrapped)) +
  geom_vline(xintercept = 1, linetype = 2) +
  geom_point(aes(size = Count, color = p.adjust), alpha = 0.9) +
  facet_wrap(~category, scales = "free_y") +
  scale_y_discrete(labels = function(x) sub("^.*?:\\s*", "", x)) +
  theme_pubr(legend = "right") +
  labs(
    x = "Fold enrichment (GeneRatio / BgRatio)",
    y = "",
    color = "FDR (BH)",
    size = "Genes in term"
  )+
  theme(axis.text.y = element_text(size = 9.5))
enrichment_plot

```

## Summary: number of genes and significant GO terms
```{r go_summary}
lost_groups <- c("not in Superscript IV 50 °C", "not in Superscript IV 42 °C", "not in Maxima H- 50 °C")

# Helper function to count significant terms
n_sig_terms <- function(enrich_obj, q = 0.05){
  if (is.null(enrich_obj)) return(0L)
  df <- as.data.frame(enrich_obj)
  if (nrow(df) == 0) return(0L)
  sum(df$p.adjust < q, na.rm = TRUE)
}

overview <- tibble(
  group = factor(lost_groups, levels = lost_groups),
  n_genes = sapply(lost_groups, \(g) length(unique(lost_lists_ez[[g]]))),
  n_go_sig = sapply(lost_groups, \(g) n_sig_terms(ego[[g]], q = 0.05))
)

# Plot 1: number of genes
p_genes <- ggplot(overview, aes(x = group, y = n_genes)) +
  geom_col(width = 0.7) +
  coord_flip() +
  theme_pubr() +
  labs(x = "", y = "# genes for GO analysis")

# Plot 2: number of significant GO terms
p_go <- ggplot(overview, aes(x = group, y = n_go_sig)) +
  geom_col(width = 0.7) +
  coord_flip() +
  theme_pubr() +
  labs(x = "", y = "# significant GO BP terms \n(FDR < 0.05)")+
  theme( axis.text.y = element_blank(),
         axis.ticks.y = element_blank())
p_genes | p_go


```
## Combine GO enrichment plots

```{r combine_go_plots}
plot_comb2 <- (p_genes | p_go )/enrichment_plot+ 
  plot_layout(
    heights = c(1, 3.5))

plot_comb2
```

## Create final combined figure

```{r final_figure}
# Define layout for combined figure
layout_RT_rev  <- c("AB\nCD\nEE")

plot_comb_all <- 
  plot_expression + 
  plot_reproducibility + 
  p_genes+
  p_go + 
  enrichment_plot+
  plot_layout(
    design = layout_RT_rev,
    heights = c(0.9, 0.5, 1)
  ) +
  plot_annotation(tag_levels = 'A')

plot_comb_all

ggsave(plot_comb_all,
       filename=here("figures/fig_S9_RT_rev.pdf"),
       height=7,
       width=13)
```

